Caz VanDevere
Week 3 Exercises
599B Data Visualization
January 23, 2019

```{r}
#install libraries
install.packages("haven")
```


```{r}
#call libraries:
library(haven)
library(ggplot2)
```

```{r}
#call datasets:
link="https://github.com/cazvan/VisualAnalytics_2_tabularData/raw/master/Small%20Dataset-Caz/policylong.dta"
link2='https://github.com/cazvan/VisualAnalytics_2_tabularData/raw/master/Small%20Dataset-Caz/policywide.dta'
link3='https://github.com/cazvan/VisualAnalytics_2_tabularData/raw/master/Small%20Dataset-Caz/statedata.dta'
link4='https://dl.dropboxusercontent.com/s/xhs2utpj1wntmzt/BRFSSFinalSup.dta'
brfssdata=read_dta(link4)
policylong=read_dta(link)
policywide=read_dta(link2)
statedata=read_dta(link3)
brfssdata=read_dta(link4)
```

```{r}
income=brfssdata[brfssdata$income]
inc=table(income)
```

```{r}

```


```{r}
#Get stuff set up:
plotthis=table(poliWA$ex)
table1=as.data.frame(plotthis)
names(table1)=c("ex","years")
table1

#make the plot:
base = ggplot(table1, aes(x=ex,y=years,color='red')) 
lolliplot=base + geom_segment(aes(y = 0,
                                  x = ex,
                                  xend=ex,
                                  yend=years),
                              ) 
lolliplot + geom_point()
```

```{r}
# BARPLOT EXERCISE #1
title2="Interview Month"
source2="BRFSS Survey Data 2000-2018"
color=rainbow(length(ToPlot)) # customizing set of colors

year=table(brfssdata$year)
barplot(year,col=color,main=title2,sub=source2,ylab='Number of Interviews',xlab="Year",)
```

```{r}
titletext="Total years since 2000 that Washington has had expanded Medicaid"
sourcetext="Source: Kaiser Family Foundation"
#" Status of State Action on the Medicaid Expansion Decision. November 26, 2018."
#https://www.kff.org/health-reform/state-indicator/state-activity-around-expanding-medicaid-under-the-af fordable-care-act/?currentTimeframe=0&sortModel=%7B%22colId%22:%22Location%22,%22sort%22:%22as c%22%7D

# WIP yearex=policylong[policylong]
#barplot(table(poliWA$ex))

poliWA=policylong[policylong$stname=="Washington",]
waex=table(poliWA$ex)

color=rainbow(length(ToPlot)) # customizing set of colors
names(waex)=c("No expansion","Expansion")
barplot(waex,
         col = color,
         main=titletext,
         sub=sourcetext,
        ylim=c(0,18),
        ylab = '(number of years)',
        )
legend(1.6, 14,
  legend = c("Did not expand ACA", "Expanded ACA"),
  fill=color,
)
```



#Notes/experiments:
#dimensions
dim(policylong)

str(policylong)

table(policylong$state)

# relative values
emrate=table(statedata$employedrate,exclude = 'nothing')
prop.table(emrate)

# relative values
stname=table(policylong$stname,exclude = 'nothing')
prop.table(stname)

# Cool super beautiful pie chart:
pieplot=prop.table(stname)
pie(pieplot)

zero=policylong[policylong$year=="2000",]
one=policylong[policylong$year=="2001",]
two=policylong[policylong$year=="2002",]
three=policylong[policylong$year=="2003",]
four=policylong[policylong$year=="2004",]
five=policylong[policylong$year=="2005",]
six=policylong[policylong$year=="2006",]
seven=policylong[policylong$year=="2007",]
eight=policylong[policylong$year=="2008",]
nine=policylong[policylong$year=="2009",]
ten=policylong[policylong$year=="2010",]
eleven=policylong[policylong$year=="2011",]
twelve=policylong[policylong$year=="2012",]
thirteen=policylong[policylong$year=="2013",]
fourteen=policylong[policylong$year=="2014",]

currentyear=policylong[policylong$year=="2018",]

#this works for older stat files
library(foreign)
policylong=read.dta(link)

# For opening csv files:
mydata=read.csv(link)



